A Theory for Data Migration of Information Systems

### 2022-10-02 2022 Migration/articleMigration.tex
minor
Todo's voor Stef en mij
Isa is a rule

@@

   Concepts are names we use during the design of a dataset
   to name a group of atoms of the same type,
   to assign signatures to relations, and thus
   to enable type checking of the dataset.
   For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
   In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
    We will use variables $A$, $B$, $C$, $D$ to represent concepts.
   The expression $a\ \inst\ A$ means that atom $a$ is an \emph{instance} of concept $A$.
   The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
   We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
   Specialization is needed to allow statements such as: ``A woman is a human'' or ``A human is a mammal''.
   The set of concepts is called $\concepts$.

@@

   Relations in datasets are used to store data.
   In this paper relations are represented by variables $r$, $s$, and $d$.
   Every relation $r$ contains a set of pairs, which we call the population of $r$:
\begin{equation}
   \pop{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\triples\}
\end{equation}
   The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.

@@

   The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
   The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
   It contains all pairs that can be made of elements of $A$ and $B$:
\[\pop{\fullt{A}{B}} = \{ \pair{a}{b}| a\ \inst\ A,\ b\ \inst\ B \}\]
   The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
   The set of relations is called $\rels$.

   We can now define the notion of dataset.
   Let   $\concepts$ be a set of concepts,
   let $\isa$ be a partial order on $\concepts$,
   let $\rels$ be a set of relations,
   let $\atoms$ be a set of atoms,
   and let $\inst$ be a relation $\atoms\times\concepts$ that represents the instance relation between concepts and atoms,
\begin{definition}[dataset]
   \label{def:dataset}
   \item A dataset is a tuple $\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$ with
   \begin{eqnarray}
      \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
      %\forall A,B\in\concepts, a\in\atoms&:&A\isa B\wedge a\ \inst\ A\ \Rightarrow\ a\ \inst\ B\\
      \forall \declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\\
      \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&\begin{array}[t]{@{}l}\declare{n}{A}{B}\in\rels\ \wedge\\ a\ (\inst \compose \kleenestar{\flip{\isa}})\ A\ \wedge\\ b\ (\inst \compose \kleenestar{\flip{\isa}})\ B\end{array}
   \end{eqnarray}
\end{definition}

\subsection{Rules}
\label{sct:Rules}
   Rules are used to restrict the populations of a dataset.
\begin{definition}[rule]
   \label{def:rule}
   \item A rule $u$ is a tuple $\la p_u,A,B,\id{viol}_u\ra$
   in which $p_u$ is a constraint (i.e. a condition that constrains the triples of a dataset),
   $\pair{A}{B}$ is the signature of $u$, and
   $\id{viol}_u:\powerset{\triples}\rightarrow\powerset{\atoms\times\atoms}$ is a function that gives the violations (which are pairs) of this rule,
   in which
\begin{eqnarray}
   \pair{a}{b}\in\viol{u}{\triples}&\Rightarrow&a\in A\wedge b\in B\\
   p_u&\Leftrightarrow&\viol{u}{\triples}=\emptyset
\end{eqnarray}
\end{definition}
   The function $\id{viol}_u$ computes the violations of $u$ for any set of triples $\triples$ in the dataset,
   which allows us to check at any moment in time whether the current population is free of violations.
   We say that a rule $u$ is \define{satisfied} in a dataset $\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$ iff $\viol{u}{\triples}=\emptyset$.
   Since $\triples$ changes continually in an information system, the violations are changing along.
   As a result, rules that are violated at one moment may be satisfied at another moment.

   The set of triples changes due to events observed by the dataset.
   These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples.
   Every rule is assigned to a role.
   That role must ensure to satisfy that rule after some event has caused violations by changing the set of triples.
   In other words: If anyone edits in the dataset and breaks a rule, someone or something has to fix the dataset to satisfy that rule again.
   Example: if a rule says that every request must be acknowledged, the receipt of a request will break this rule.
   Sending an acknowledgement message will cause the rule to be satisfied again.

\subsection{Information Systems}
\label{sct:Information Systems}
   The purpose of an information system is to make data meaningful to users.
   Data migration is all about preserving that meaning.
   This section introduces the notion of information system
   as a system that consist of a dataset and a set of rules that constrain that data,
   together with a set of roles that must keep all rules satisfied.
   These rules reflect the meaning (semantics) that is shared among users.

@@

   A \define{role} is a name that identifies a group of users.
   The purpose of a role is to mention an individual user without knowing who that user is.
   In the sequel, when we talk about a role we actually mean a user (an arbitrary one) who fulfills that role in the information system.
   The set of roles is called $\roles$.

   We can now define the notion of information system.
   Let $\roles$ be a set of roles,
   let $\rules$ be a set of rules, 
   let $\maintain$ be a surjective relation $\roles\times\rules$,
   and let $\dataset$ be a dataset.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\la\roles,\rules,\maintain,\dataset\ra$\\
   with dataset $\dataset=\la\atoms,\concepts,\inst,\isa,\rels,\triples\ra$
\end{definition}
   The relation $\maintain$ links rules to roles.
   $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{u}{\triples}$ empty.
   If a rule $u$ has violations, i.e. $\viol{u}{\triples}$ is not empty, there is a role whose job it is to remove those violations.
   The relation $\maintain$ being surjective means that all users together are keeping the violation set empty.

>>>>

% TODO Bas:
% - verhaal over incrementeel fouten oplossen (fouten staan al in artikel)

% TODO Algemeen:
% - type van triple-set en atom-set definieren zodat viol_u ?->P(?x?)
% - nieuwe populatie na de disjoint union zou de populatie met ENFORCE regels toegepast moeten zijn

% TODO Stef:
% - stukje literatuuronderzoek
% https://ieeexplore.ieee.org/abstract/document/7445334?casa_token=ECzi6XeV2ncAAAAA:KhWzB8XBFOUJ0C6AD-XjX_ryuA9ARvTd3gm6RR-ZNiR8sZ1858FJpQ7zKQhkAZDlv8IjPdgD
% https://ieeexplore.ieee.org/abstract/document/8549944?casa_token=9qiGqNzh2Q0AAAAA:C-cYogExB35nGxQdxLcdBh4JoLNvM0OHedAMhCbB5V4kb4_6nzHUvc23xSJbeoBu67LSiz-Y
% https://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.651.9298&rep=rep1&type=pdf
% https://www.scirp.org/html/4-7800724_106592.htm
% https://www.researchgate.net/profile/Ranjana-Badre/publication/318665687_GUI_for_Data_Migration_and_Query_Conversion/links/5b45bbea0f7e9b1c722386e5/GUI-for-Data-Migration-and-Query-Conversion.pdf
% https://journal3.uin-alauddin.ac.id/index.php/literatify/article/view/12567

@@

   Concepts are names we use during the design of a dataset
   to name a group of atoms of the same type,
   to assign signatures to relations, and thus
   to enable type checking of the dataset.
   For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
   In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
    We will use variables $A$, $B$, $C$, $D$ to represent concepts.
   The expression $a\ \inst\ A$ means that atom $a$ is an \emph{instance} of concept $A$.
   The set of concepts is called $\concepts$.

@@

   Relations in datasets are used to store data.
   In this paper relations are represented by variables $r$, $s$, and $d$.
   Every relation $r$ contains a set of pairs, which we call the population of $r$:
\begin{equation}
   \pop{\triples}{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\triples\}
\end{equation}
   The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.

@@

   The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
   The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
   It contains all pairs that can be made of elements of $A$ and $B$:
\[\pop{\triples}{\fullt{A}{B}} = \{ \pair{a}{b}| a\ \inst\ A,\ b\ \inst\ B \}\]
   The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
   The set of relations is called $\rels$.

   We can now define the notion of dataset.
   Let   $\concepts$ be a set of concepts,
   let $\isa$ be a partial order on $\concepts$,
   let $\rels$ be a set of relations,
   let $\atoms$ be a set of atoms,
   and let $\inst$ be a relation $\atoms\times\concepts$ that represents the instance relation between concepts and atoms,
\begin{definition}[dataset]
   \label{def:dataset}
   \item A dataset is a tuple $\la\atoms,\concepts,\inst,\rels,\triples\ra$ with
   \begin{eqnarray}
      \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\\
      %\forall A,B\in\concepts, a\in\atoms&:&A\isa B\wedge a\ \inst\ A\ \Rightarrow\ a\ \inst\ B\\
      \forall \declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\\
      \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&\begin{array}[t]{@{}l}\declare{n}{A}{B}\in\rels\ \wedge\ a\ \inst\ A\ \wedge\ b\ \inst\ B\end{array}
   \end{eqnarray}
\end{definition}

\subsection{Rules}
\label{sct:Rules}
   Rules are used to describe whether certain (assumably desirable) properties hold in an information system.

\begin{definition}[rule]
   \label{def:rule}
   \item A rule $u$ is a tuple $\la A,B,\id{viol}_u\ra$.
   %in which $p_u$ is a constraint (i.e. a condition that constrains the triples of a dataset),
   $\pair{A}{B}$ is the signature of $u$ and
   $\id{viol}_u$ is a function that gives the violations (which are pairs) of this rule,
   in which
\begin{eqnarray}
   \pair{a}{b}\in\viol{u}{\triples,\inst}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
\end{eqnarray}
\end{definition}

   The function $\id{viol}_u$ computes the violations of $u$,
   which allows us to check at any moment in time whether the current population is free of violations.
   We say that a rule $u$ is \define{satisfied} in a dataset $\la\atoms,\concepts,\inst,\rels,\triples\ra$ iff $\viol{u}{\triples}=\emptyset$.
   This allows us to define a predicate $p_u$ that indicates if a rule $u$ is satisfied:
\begin{eqnarray}
   p_u(\triples,\inst)&\Leftrightarrow&\viol{u}{\triples,\inst}=\emptyset
\end{eqnarray}

   Since $\triples$ changes continually in an information system, the violations are changing along.
   As a result, rules that are violated at one moment may be satisfied at another moment.
   The set of triples changes due to events observed by the information system.
   These events may be caused by external systems, roles acting on the dataset, or any other force that changes triples.
   Every rule is assigned to a role.
   That role must ensure to satisfy that rule after some event has caused violations by changing the set of triples.
   In other words: If anyone edits in the dataset and breaks a rule, someone or something has to fix the dataset to satisfy that rule again.
   Example: if a rule says that every request must be acknowledged, the receipt of a request will break this rule.
   Sending an acknowledgement message will cause the rule to be satisfied again.

   Ampersand supports a special kind of rule that is used for specialization.
   The rule $A\isa B = \la A, A, viol_{A\isa B} \ra$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
   Instances of $A$ that are not an instance of $B$ are the violations of this rule: $viol_{A\isa B}(\triples,\inst) = \{(a,a) \mid a\ \inst\ A \wedge \lnot (a\ \inst\ B)\}$.
   We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.

      Specialization rules are special for two reasons:
   first, $A\isa B$ is automatically maintained by the information system itself, by changing $\inst$ to include any violating pairs in $B$.
   This means that an information system can never be brought into a state in which $p_{A\isa B}$ is False if $A\isa B$ is a specialization rule of the information system, allowing for certain run-time optimizations.
   A second property that sets specialization rules apart, is that they play a role in the Ampersand compiler.
   For example, code that determines equality between atoms only works for atoms that are an instance of the same concept.
   To ensure that this does not result in problems at run-time, users are required to write specialization rules that make these comparisons possible, and the type checker in the Ampersand compiler will spot these issues if they occur and refuse to generate the information system.

\subsection{Information Systems}
\label{sct:Information Systems}
   The purpose of an information system is to make data meaningful to users.
   Data migration is all about preserving that meaning.
   This section introduces the notion of information system
   as a system that consist of a dataset and a set of rules that constrain that data,
   together with a set of roles that must keep all rules satisfied.
   %These rules reflect the meaning (semantics) that is shared among users.

@@

   A \define{role} is a name that identifies a group of users.
   The purpose of a role is to mention an individual user without knowing who that user is.
   In the sequel, when we talk about a role we actually mean a user (an arbitrary one) who fulfills that role in the information system.

   We can now define the notion of information system.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\la\roles,\rules,\specrules,\maintain,\dataset\ra$.
   Here $\roles$ is a set of roles,
   $\rules$ a set of rules,
   $\specrules$ a set of specialization rules,
   $\maintain$ is a relation $\roles\times\rules$,
   and $\dataset$ is a dataset.
\end{definition}
   The relation $\maintain$ links rules to roles.
   $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{u}{\triples}$ empty.
   If a rule $u$ has violations, i.e. $\viol{u}{\triples}$ is not empty, there is a role whose job it is to remove those violations.
   The relation $\maintain$ being surjective means that all users together are keeping the violation set empty.
